Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2024-08-08 paper/main.tex
Updates

In contrast, the SEM literature has focused more on developing tools to assist
researchers in manually constructing or modifying models while allowing them to
integrate their expert knowledge into the model-building process. Researchers
typically develop an initial model based on their expertise and then use
various tools to guide modifications to improve the model's fit to data. This
model modification process is commonly known as \emph{Specification Search}
\citep{Long1983}, employing method like modification indices, and Wald-based
tests \citep{Marcoulides2018}.

@@

\begin{enumerate}
	\item \textbf{Obvious Mistakes:} Although most algorithms are
		theoretically consistent asymptotically, their finite sample
		properties are not well understood. Moreover, their output can
		vary significantly depending on the selected algorithm and
		their hyperparameters. Combined with the lack of robust methods
		to test these algorithms' performance on a given dataset, it
		can be difficult to trust their outputs. Consequently,
		researchers have to rely on post-hoc modifications to the model
		based on their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges before using them for analysis.
\end{enumerate}

Due to these challenges, even when using automated algorithms, we often have to
rely on manual post hoc testing and modifications to the model learned by the
algorithm. Whether constructing the model manually or using an algorithm, we
must rely on domain expertise to construct or modify the model. Hence, it is
crucial to have tools to guide this modification process. One of the ways to
guide this modification is to use Conditional Independence (CI) based model
testing \cite{Ankan2023}. As DAGs imply CI statements that can be tested in the
given data using statistical CI tests, we can verify whether the implied CIs
are consistent in the data or not. If some of implied CIs are not valid in the
given dataset, we can modify the model accordingly. Figure~\ref{fig:ci_testing}
shows an example of model testing using CIs. However, using p-values of the CI
tests has a few drawbacks. Firstly, as CIs are implied only by missing edges in
the model, we can not test whether an incorrect edge is already present in the
model. For examples, in high sample size scenarios, we might get a significant
association between a pair of variables with very small association. Secondly,
the number of implied CIs can be quite large. For example, in the moderately
sized alarm network with $ 37 $ node, there are $287$ implied CIs. If we rely
only on the p-values of the CI tests, it can be difficult to figure which test
to focus on.

>>>>

In contrast, the SEM literature has focused more on developing tools to assist
researchers in manually constructing or modifying models while allowing them to
integrate their expert knowledge into the model-building process. Researchers
typically develop an initial model based on their expertise and then use
various tools to guide modifications to improve the model's fit to data. This
model construction process is commonly known as \emph{Specification Search}
\citep{Long1983} and uses method like modification indices, and Wald-based
tests \citep{Marcoulides2018}.

@@

\begin{enumerate}
	\item \textbf{Lack of Trust:} Although most algorithms are
		asymptotically consistent, their finite sample properties are
		not well understood. Moreover, their output can vary
		significantly depending on the algorithm and their
		hyperparameters. Combined with the lack of robust methods to
		test these algorithms' performance on a given dataset, it can
		be difficult to trust their outputs. Consequently, researchers
		usually have to rely on post-hoc modifications to the model
		based on their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges before using them for analysis.
\end{enumerate}

Figure shows some of these issues when applied to the Adult Income dataset. Due
to these challenges, even when using automated algorithms, we often have to
rely on manual post hoc testing and modifications to the model learned by the
algorithm. Whether constructing the model manually or using an algorithm, we
must rely on domain expertise to construct or modify the model. Hence, it is
crucial to have tools to guide this modification process. One of the ways to
guide this modification is to use Conditional Independence (CI) based model
testing \cite{Ankan2023}. As DAGs imply CI statements that can be tested in the
given data using statistical CI tests, we can verify whether the implied CIs
are consistent in the data or not. If some of implied CIs are not valid in the
given dataset, we can modify the model accordingly. Figure~\ref{fig:ci_testing}
shows an example of model testing using CIs. However, using p-values of the CI
tests has a few drawbacks. Firstly, as CIs are implied only by missing edges in
the model, we can not test whether an incorrect edge is already present in the
model. For examples, in high sample size scenarios, we might get a significant
association between a pair of variables with very small association. Secondly,
the number of implied CIs can be quite large. For example, in the moderately
sized alarm network with $ 37 $ node, there are $287$ implied CIs. If we rely
only on the p-values of the CI tests, it can be difficult to figure which test
to focus on.
